Lesson 6
========================================================

### Welcome
Notes:
```{r}
library('ggplot2')
data(diamonds)
```
***

### Scatterplot Review

```{r Scatterplot Review}
# Let's start by examining two variables in the data set.
# The scatterplot is a powerful tool to help you understand
# the relationship between two continuous variables.

# We can quickly see if the relationship is linear or not.
# In this case, we can use a variety of diamond
# characteristics to help us figure out whether
# the price advertised for any given diamond is 
# reasonable or a rip-off.

# Let's consider the price of a diamond and it's carat weight.
# Create a scatterplot of price (y) vs carat weight (x).

# Limit the x-axis and y-axis to omit the top 1% of values.

ggplot(aes(x = carat, y = price),
       data = diamonds) +
  geom_point(fill = I('#F79420'), color = I('black'), shape = 21) + # params from video
  scale_x_continuous(limits = c(0, quantile(diamonds$carat, 0.99))) +
  scale_y_continuous(limits = c(0, quantile(diamonds$price, 0.99)))
# video sloution
qplot(data = diamonds, x = carat, y = price,
       xlim = c(0, quantile(diamonds$carat, 0.99)),
       ylim = c(0, quantile(diamonds$price, 0.99))) +
  geom_point(fill = I('#F79420'), color = I('black'), shape = 21)
```

***

### Price and Carat Relationship
Response:
What do you notice about the relationship between price and carat?

In general, the price increases as carat weight increases, but is seems like price increases faster as carat size increases (it looks like an exponential increase).

dispersion or variance increases as carat size increases.
```{r}
# add linear trendline
ggplot(aes(x = carat, y = price), data = diamonds) +
  geom_point(fill = I('#F79420'), color = I('black'), shape = 21) +
  stat_smooth(method = 'lm') +
  scale_x_continuous(limits = c(0, quantile(diamonds$carat, 0.99))) +
  scale_y_continuous(limits = c(0, quantile(diamonds$price, 0.99)))
# notice that trendline doesn't pass through center of data at some key places  
```
***

### Frances Gerety
Notes:
Frances Gerety coined a famous slogan.
Complete that slogan.
  "A diamond is...
          forever."
  
Have You Ever Tried to Sell a Diamond?:
https://www.theatlantic.com/magazine/archive/1982/02/have-you-ever-tried-to-sell-a-diamond/304575/

De Beers' Diamond Monopoly:
https://en.wikipedia.org/wiki/De_Beers#Diamond_monopoly
#### A diamonds is
forever.
***

### The Rise of Diamonds
Notes:
1980s ad introduced 2 months salary benchmark.
***

### ggpairs Function
Notes:

```{r ggpairs Function}
# install these if necessary
install.packages('GGally')
install.packages('scales')
install.packages('memisc')
install.packages('lattice')
install.packages('MASS')
install.packages('car')
install.packages('reshape')
install.packages('plyr')
```

```{r ggpairs Function}
# load the ggplot graphics package and the others
library(ggplot2)
library(GGally)
library(scales)
library(memisc)

# sample 10,000 diamonds from the data set
set.seed(20022012)
diamond_samp <- diamonds[sample(1:length(diamonds$price), 10000), ]
# comment out deprecated syntax
# ggpairs(diamond_samp, params = c(shape = I('.'), outlier.shape = I('.')))
ggpairs(diamond_samp,
  lower = list(continuous = wrap("points", shape = I('.'))),
  upper = list(combo = wrap("box", outlier.shape = I('.'))))

```

What are some things you notice in the ggpairs output?
Response:

***

### The Demand of Diamonds
Notes:

```{r The Demand of Diamonds}
# Create two histograms of the price variable
# and place them side by side on one output image.

# We’ve put some code below to get you started.

# The first plot should be a histogram of price
# and the second plot should transform
# the price variable using log10.

# Set appropriate bin widths for each plot.
# ggtitle() will add a title to each histogram.

# You can self-assess your work with the plots
# in the solution video.

# ALTER THE CODE BELOW THIS LINE
# ==============================================

library(gridExtra)

plot1 <- qplot(x = price, data = diamonds, binwidth = 100) + 
  ggtitle('Price')

plot2 <- qplot(x = price, data = diamonds, binwidth = 0.01) +
  ggtitle('Price (log10)') +
  scale_x_log10()

grid.arrange(plot1, plot2, nrow = 1)
```

Log Transformations:
http://www.r-statistics.com/2013/05/log-transformations-for-skewed-and-wide-distributions-from-practical-data-science-with-r/
***

### Connecting Demand and Price Distributions
Notes:
When looking at these plots, what do you notice? Think specifically about the two peaks in the transformed plot and how it relates to demand for diamonds.

The log transformation shows a bimodal distribution, where there's a higher peak around a certain high price range. Therefore, there seems to be separate demand peaks in the low and upper ranges. This is consistent with the rich and poor buyer theory from the previous exploration.
***

### Scatterplot Transformation

```{r Scatterplot Transformation}
qplot(carat, price, data = diamonds) +
  scale_y_continuous(trans = log10_trans() ) +
  ggtitle('Price (log10) by Carat')
```
This is looking better, but we can do better still by remembering that carat is a function of volume that is also a function of 3 dimensions. So a cuberoot of carat, might be worth examining.

### Create a new function to transform the carat variable

```{r cuberoot transformation}
# volume is a cube function of the three dmensions
cuberoot_trans = function() trans_new('cuberoot', transform = function(x) x^(1/3),
                                      inverse = function(x) x^3)
```

#### Use the cuberoot_trans function
```{r Use cuberoot_trans}
ggplot(aes(carat, price), data = diamonds) + 
  geom_point() + 
  scale_x_continuous(trans = cuberoot_trans(), limits = c(0.2, 3),
                     breaks = c(0.2, 0.5, 1, 2, 3)) + 
  scale_y_continuous(trans = log10_trans(), limits = c(350, 15000),
                     breaks = c(350, 1000, 5000, 10000, 15000)) +
  ggtitle('Price (log10) by Cube-Root of Carat')
```
after doing the transformation, the graph looks almost linear.
***

### Overplotting Revisited

```{r Sort and Head Tables}
head(sort(table(diamonds$carat), decreasing = T))
head(sort(table(diamonds$price), decreasing = T))
```
can compensate by making points smaller, adding jitter, or adding transparency.

```{r Overplotting Revisited}
# Add a layer to adjust the features of the
# scatterplot. Set the transparency to one half,
# the size to three-fourths, and jitter the points.

# If you need hints, see the Instructor Notes.
# There are three hints so scroll down slowly if
# you don’t want all the hints at once.

# ALTER THE CODE BELOW THIS LINE
# =======================================================================

# ggplot(aes(carat, price), data = diamonds) + 
#   geom_point() + 
#   scale_x_continuous(trans = cuberoot_trans(), limits = c(0.2, 3),
#                      breaks = c(0.2, 0.5, 1, 2, 3)) + 
#   scale_y_continuous(trans = log10_trans(), limits = c(350, 15000),
#                      breaks = c(350, 1000, 5000, 10000, 15000)) +
#   ggtitle('Price (log10) by Cube-Root of Carat')

ggplot(aes(carat, price), data = diamonds) + 
  geom_point(alpha = 1/2, size = 3/4, position = "jitter") + 
  scale_x_continuous(trans = cuberoot_trans(), limits = c(0.2, 3),
                     breaks = c(0.2, 0.5, 1, 2, 3)) + 
  scale_y_continuous(trans = log10_trans(), limits = c(350, 15000),
                     breaks = c(350, 1000, 5000, 10000, 15000)) +
  ggtitle('Price (log10) by Cube-Root of Carat')
```

***

### Other Qualitative Factors
Notes:
What makes a diamond sparkle?
http://www.bluenile.com/diamonds/diamond-cut
Does clarity affect the sparkle? 
http://www.bluenile.com/diamonds/diamond-clarity
What's a jeweler's loupe?
http://en.wikipedia.org/wiki/Loupe
***

### Price vs. Carat and Clarity

Alter the code below.
```{r Price vs. Carat and Clarity}
# Adjust the code below to color the points by clarity.

# A layer called scale_color_brewer() has 
# been added to adjust the legend and
# provide custom colors.

# See if you can figure out what it does.
# Links to resources are in the Instructor Notes.

# You will need to install the package RColorBrewer
# in R to get the same colors and color palettes.

# =======================================
# install and load the RColorBrewer package
# install.packages('RColorBrewer')
library(RColorBrewer)

# ggplot(aes(x = carat, y = price), data = diamonds) + 
#   geom_point(alpha = 0.5, size = 1, position = 'jitter') +
#   scale_color_brewer(type = 'div',
#     guide = guide_legend(title = 'Clarity', reverse = T,
#     override.aes = list(alpha = 1, size = 2))) +  
#   scale_x_continuous(trans = cuberoot_trans(), limits = c(0.2, 3),
#     breaks = c(0.2, 0.5, 1, 2, 3)) + 
#   scale_y_continuous(trans = log10_trans(), limits = c(350, 15000),
#     breaks = c(350, 1000, 5000, 10000, 15000)) +
#   ggtitle('Price (log10) by Cube-Root of Carat and Clarity')

ggplot(aes(x = carat, y = price, colour = clarity), data = diamonds) + 
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  scale_color_brewer(type = 'div',
    guide = guide_legend(title = 'Clarity', reverse = T,
    override.aes = list(alpha = 1, size = 2))) +  
  scale_x_continuous(trans = cuberoot_trans(), limits = c(0.2, 3),
    breaks = c(0.2, 0.5, 1, 2, 3)) + 
  scale_y_continuous(trans = log10_trans(), limits = c(350, 15000),
    breaks = c(350, 1000, 5000, 10000, 15000)) +
  ggtitle('Price (log10) by Cube-Root of Carat and Clarity')
```
ggplot2: scale_colour_brewer
http://docs.ggplot2.org/current/scale_brewer.html
ggplot2: Color Brewer Palettes and Safe Colors
http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/#palettes-color-brewer
ggplot2: Legends
http://www.cookbook-r.com/Graphs/Legends_(ggplot2)/
***

### Clarity and Price
Response:
Based on the plot, do you think clarity explains some of the change in price? Why?

Yes, clarity explains some of the change in price. There are clear bands of clarity for all carat weights, so for any size there a change in price based on clarity.
***

### Price vs. Carat and Cut

Alter the code below.
```{r Price vs. Carat and Cut}
# Let’s look at cut and see if we find a similar result.

# Adjust the code below to color the points by cut.
# Change any other parts of the code as needed.

# ALTER THE CODE BELOW THIS LINE
# =========================================================================================
ggplot(aes(x = carat, y = price, color = cut), data = diamonds) + 
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = 'Cut', reverse = T,
                                          override.aes = list(alpha = 1, size = 2))) +  
  scale_x_continuous(trans = cuberoot_trans(), limits = c(0.2, 3),
                     breaks = c(0.2, 0.5, 1, 2, 3)) + 
  scale_y_continuous(trans = log10_trans(), limits = c(350, 15000),
                     breaks = c(350, 1000, 5000, 10000, 15000)) +
  ggtitle('Price (log10) by Cube-Root of Carat and Cut')
```

***

### Cut and Price
Response:
Based on the plot, do you think that cut accounts for some of the variance in price? Why?

No. The cuts are overlapping for all carat weights and prices. Therefore, cut doesn't seem to have a clear affect on price.

What makes a diamond sparkle?
http://www.bluenile.com/diamonds/diamond-cut
***

### Price vs. Carat and Color

Alter the code below.
```{r Price vs. Carat and Color}
# Finally, let’s use diamond color to color our plot.

# Adjust the code below to color the points by diamond colors
# and change the titles.

# ALTER THE CODE BELOW THIS LINE
# ===========================================================================================
ggplot(aes(x = carat, y = price, color = color), data = diamonds) + 
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = 'Color', reverse = FALSE,
                                          override.aes = list(alpha = 1, size = 2))) +  
  scale_x_continuous(trans = cuberoot_trans(), limits = c(0.2, 3),
                     breaks = c(0.2, 0.5, 1, 2, 3)) + 
  scale_y_continuous(trans = log10_trans(), limits = c(350, 15000),
                     breaks = c(350, 1000, 5000, 10000, 15000)) +
  ggtitle('Price (log10) by Cube-Root of Carat and Color')
```

***

### Color and Price
Response:
Based on the plot, do you think that the color of a diamond influences price? Why?

Yes, there are consistent price stratifications based on color for all carat weights.

Diamond Education: Color
http://www.bluenile.com/diamonds/diamond-color
***

### Linear Models in R
Notes:
use lm() function. supply function in the form of lm(y~x)
Response:

log(price)~carat^(1/3)

Nice! Price is the outcome and carat is the predictor variable. We used our domain knowledge of diamonds and carat weight to take the cube root of carat weight (volume).
***

### Building the Linear Model
Notes:
I stands for "as is"
```{r Building the Linear Model}
m1 <- lm(I(log(price)) ~ I(carat^(1/3)), data = diamonds)
m2 <- update(m1, ~ . + carat)
m3 <- update(m2, ~ . + cut)
m4 <- update(m3, ~ . + color)
m5 <- update(m4, ~ . + clarity)
mtable(m1, m2, m3, m4, m5, sdigits = 3)
```

Notice how adding cut to our model does not help explain much of the variance
in the price of diamonds. This fits with out exploration earlier.

Linear Models and Operators in R
http://data.princeton.edu/R/linearModels.html
***

### Model Problems
Video Notes:
What could be some problems when using this model? What else should we think about when using this model?
Research:
(Take some time to come up with 2-4 problems for the model)
(You should 10-20 min on this)

Response:
This model doesn't take changing prices over time into account due to asset appreciation. Also, sharp changes in demand due to market (ad campaign or boycott) or supply by closing a mine or tariff would have a drastic change to the price.

video:
from 2008
- inflation
- 2008 global financial recession (prices dropped)
- diamond market in china heating up
- uneven recovery/price increase across different carat weight
***

### A Bigger, Better Data Set
Notes:

```{r A Bigger, Better Data Set}
# install.package('bitops')
# install.packages('RCurl')
library('bitops')
library('RCurl')

diamondsurl = getBinaryURL("https://raw.github.com/solomonm/diamonds-data/master/BigDiamonds.Rda")
load(rawConnection(diamondsurl))
```
The code used to obtain the data is available here:
https://github.com/solomonm/diamonds-data

```{r}
# Your task is to build five linear models like Solomon
# did for the diamonds data set only this
# time you'll use a sample of diamonds from the
# diamondsbig data set.

# Be sure to make use of the same variables
# (logprice, carat, etc.) and model
# names (m1, m2, m3, m4, m5).

# To get the diamondsbig data into RStudio
# on your machine, copy, paste, and run the
# code in the Instructor Notes. There's
# 598,024 diamonds in this data set!

# Since the data set is so large,
# you are going to use a sample of the
# data set to compute the models. You can use
# the entire data set on your machine which
# will produce slightly different coefficients
# and statistics for the models.

# This exercise WILL BE automatically graded.

# You can leave off the code to load in the data.
# We've sampled the data for you.
# You also don't need code to create the table output of the models.
# We'll do that for you and check your model summaries (R^2 values, AIC, etc.)

# Your task is to write the code to create the models.

# DO NOT ALTER THE CODE BELOW THIS LINE (Reads in a sample of the diamondsbig data set)
#===========================================================================================
# diamondsBigSample <- read.csv('diamondsBigSample.csv')
diamondsbig <- read.csv('diamondsbig.csv')

# ENTER YOUR CODE BELOW THIS LINE. (Create the five models)
#===========================================================================================
# m1 <- lm(I(log(price)) ~ I(carat^(1/3)), data = diamondsBigSample)
# m2 <- update(m1, ~ . + carat)
# m3 <- update(m2, ~ . + cut)
# m4 <- update(m3, ~ . + color)
# m5 <- update(m4, ~ . + clarity)
# mtable(m1, m2, m3, m4, m5, sdigits = 3)

# video version:
diamondsbig$logprice = log(diamondsbig$price)
m1 <- lm(logprice ~ I(carat^(1/3)),
         data=diamondsbig[diamondsbig$price < 10000 &
                            diamondsbig$cert == "GIA",])
m2 <- update(m1, ~ . + carat)
m3 <- update(m2, ~ . + cut)
m4 <- update(m3, ~ . + color)
m5 <- update(m4, ~ . + clarity)
```
## Building a Model Using the Big Diamonds Data Set
Notes:

```{r Building a Model Using the Big Diamonds Data Set}

```


***

## Predictions

Example Diamond from BlueNile:
Round 1.00 Very Good I VS1 $5,601

```{r}
#Be sure you’ve loaded the library memisc and have m5 saved as an object in your workspace.
thisDiamond = data.frame(carat = 1.00, cut = "V.Good",
                         color = "I", clarity="VS1")
modelEstimate = predict(m5, newdata = thisDiamond,
                        interval="prediction", level = .95)
```

Evaluate how well the model predicts the BlueNile diamond's price. Think about the fitted point estimate as well as the 95% CI.

# It estimated it pretty well, but the confidence interval is very large, so there isn't good certainty for the predictions throughout the model.

The prediction interval here may be slightly conservative, as the model errors are heteroskedastic over carat (and hence price) even after our log and cube-root transformations.

See the output of the following code.
```{r}
dat = data.frame(m4$model, m4$residuals)

with(dat, sd(m4.residuals))

with(subset(dat, carat > .9 & carat < 1.1), sd(m4.residuals))

dat$resid <- as.numeric(dat$m4.residuals)
ggplot(aes(y = resid, x = round(carat, 2)), data = dat) +
  geom_line(stat = "summary", fun.y = sd)
```
How could we do better? If we care most about diamonds with carat weights between 0.50 and 1.50, we might restrict the data we use to fit our model to diamonds that are that size - we have enough data.
***

## Final Thoughts
Notes:
Tiffany vs. Costco: Which Diamond Ring is Better 
http://www.businessweek.com/articles/2013-05-06/tiffany-vs-dot-costco-which-diamond-ring-is-better
But Costco Sells Pricy Diamonds Too
http://www.costco.com/CatalogSearch?catalogId=10701&langId=-1&keyword=Engagement&storeId=10301&refine=30108%2b209%2b238%2b210%2b


Learn how to analyze your Facebook social network and more!
http://blog.revolutionanalytics.com/2013/11/how-to-analyze-you-facebook-friends-network-with-r.html
Thank you for joining us! Please fill out this brief survey about the course to let us know how we're doing and what you'd like to see in the future!
***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

