---
title: "Lesson 4"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r Load 'diamonds'}
library(ggplot2) #must load the ggplot package first
data('diamonds') #loads the diamonds data set since it comes with the ggplot package

# get some basic information about dataset
names(diamonds)
str(diamonds)
levels(diamonds$color)
?diamonds
```

```{r Price Histogram}
# Create a histogram of the price of
# all the diamonds in the diamond data set.
qplot(x = price, data = diamonds)
```

```{r Price Histogram Summary}
summary(diamonds$price)
```

```{r Diamond Counts}
# How many diamonds cost less than $500?
nrow(subset(diamonds, diamonds$price < 500))
# How many diamonds cost less than $250?
nrow(subset(diamonds, diamonds$price < 250))
# How many diamonds cost $15,000 or more?
nrow(subset(diamonds, diamonds$price >= 15000))
```

```{r Cheaper Diamonds}
qplot(x = price, data = subset(diamonds, !is.na(diamonds$price)), binwidth = 100,
      color = I('black'), fill = I('#099DD9')) +
  scale_x_continuous(limits=c(300,10000))

# breaks=c(0,500,10)
qplot(x = price, data = subset(diamonds, !is.na(diamonds$price)),
      color = I('black'), fill = I('#099DD9')) +
  scale_x_log10(limits=c(300,10000))

qplot(x = price, data = subset(diamonds, !is.na(diamonds$price)), binwidth = 100,
      color = I('black'), fill = I('#099DD9')) +
  scale_x_continuous(limits=c(300,2500))
```

```{r Price by Cut - Histograms}
# Break out the histogram of diamond prices by cut.
qplot(x = price,
      data = subset(diamonds, !is.na(diamonds$price)),
      binwidth = 100,
      color = cut) +
  scale_x_continuous(limits=c(300,2500))

# You should have five histograms in separate
# panels on your resulting plot.
qplot(x = price,
      data = subset(diamonds, !is.na(diamonds$price)),
      binwidth = 100) +
  scale_x_continuous(limits=c(300,2500)) +
  facet_grid(cut ~.)
```
```{r Price by Cut - 1}
# Which cut has the highest priced diamond?
by(diamonds$price, diamonds$cut, max)
```
```{r Price by Cut - 2}
# Which cut has the lowest priced diamond?
by(diamonds$price, diamonds$cut, min)
```
```{r Price by Cut - 3}
# Which cut has the lowest median price?
by(diamonds$price, diamonds$cut, median)
```

```{r Scales and Multiple Histograms}
qplot(x = price, data = diamonds) + facet_wrap(~cut, scales="free_y", ncol=1)
```
```{r Price per Carat by Cut}
# Create a histogram of price per carat
# and facet it by cut. You can make adjustments
# to the code from the previous exercise to get
# started.

# Adjust the bin width and transform the scale
# of the x-axis using log10
qplot(x = price/carat,
      data = diamonds,
      color = cut) +
  scale_x_log10()

qplot(x = price/carat,
      data = diamonds,
      bins = 20) +
  facet_wrap(~cut, nrow = 1) +
  scale_x_log10()

qplot(x = price/carat,
      data = diamonds,
      bins = 20) +
  scale_x_log10() +
  facet_wrap(~cut, nrow = 1)
```
```{r Price Box Plots - Visualizations}
# Investigate the price of diamonds using box plots,
# numerical summaries, and one of the following categorical
# variables: cut, clarity, or color.

# There won’t be a solution video for this
# exercise so go to the discussion thread for either
# BOXPLOTS BY CLARITY, BOXPLOT BY COLOR, or BOXPLOTS BY CUT
# to share you thoughts and to
# see what other people found.

# load library to make grid
library(gridExtra)

# define individual histgram plots
p1 <- qplot(x = price/carat,
      data = diamonds,
      bins = 20) +
  scale_x_log10()
p2 <- p1 + facet_wrap(~clarity, nrow = 1)
p3 <- p1 + facet_wrap(~color, nrow = 1)
p4 <- p1 + facet_wrap(~cut, nrow = 1)
# arrange plots in grid
grid.arrange(p2, p3, p4, nrow=3)


# # define individual box plots
b2 <- qplot(x = clarity, y = price/carat,
            data = diamonds, geom = 'boxplot') +
      coord_cartesian(ylim = c(2000,6000))
b3 <- qplot(x = color, y = price/carat,
            data = diamonds, geom = 'boxplot') +
      coord_cartesian(ylim = c(2000,6000))
b4 <- qplot(x = cut, y = price/carat,
            data = diamonds, geom = 'boxplot') +
      coord_cartesian(ylim = c(2000,6000))
# arrange plots in grid
grid.arrange(b2, b3, b4, nrow=3)
```
```{r Price Box Plots - Sumary Statistics}
"clarity"
by(diamonds$price/diamonds$carat, diamonds$clarity, summary)
"color"
by(diamonds$price/diamonds$carat, diamonds$color, summary)
"cut"
by(diamonds$price/diamonds$carat, diamonds$cut, summary)
```
```{r Interquartile Range - IQR}
"IQR of price for color = D"
Q1_D = quantile(diamonds$price[diamonds$color == 'D'], 0.25)
Q3_D = quantile(diamonds$price[diamonds$color == 'D'], 0.75)
IQR_D = IQR(diamonds$price[diamonds$color == 'D'])
names(IQR_D) <- c('IQR')
print(c(Q1_D, Q3_D, IQR_D))

"IQR of price for color = J"
Q1_J = quantile(diamonds$price[diamonds$color == 'J'], 0.25)
Q3_J = quantile(diamonds$price[diamonds$color == 'J'], 0.75)
IQR_J = IQR(diamonds$price[diamonds$color == 'J'])
names(IQR_J) <- c('IQR')
print(c(Q1_J, Q3_J, IQR_J))
# IQR for diamons with best and worst color
#
# according to ?diamonds:
#   diamond colour, from J (worst) to D (best)
```
```{r Interquartile Range - IQR (another way)}
price_by_color = by(diamonds$price, diamonds$color, summary)
#print(price_by_color)

"IQR of price for color = D"
Q1_D = price_by_color$D[2]
Q3_D = price_by_color$D[5]
IQR_D = Q3_D - Q1_D
names(IQR_D) <- c('IQR')
print(c(Q1_D, Q3_D, IQR_D))

"IQR of price for color = J"
Q1_J = price_by_color$J[2]
Q3_J = price_by_color$J[5]
IQR_J = Q3_J - Q1_J
names(IQR_J) <- c('IQR')
print(c(Q1_J, Q3_J, IQR_J))
```
```{r Price per Carat Box Plots by Color}
# Investigate the price per carat of diamonds across
# the different colors of diamonds using boxplots.
# # define individual box plots
qplot(x = color, y = price/carat,
          data = diamonds, geom = 'boxplot') +
      coord_cartesian(ylim = c(2400,5500))
```
```{r Carat Frequency Polygon}
qplot(x = carat, data = subset(diamonds, !is.na(carat)),
      binwidth = 0.1, geom = 'freqpoly') + 
      scale_x_continuous(breaks = seq(0,5,by=0.1), limits = c(0,3)) +
      scale_y_continuous(breaks = seq(0,15000,by=1000))
```

```{r Gapminder Data}
# The Gapminder website contains over 500 data sets with information about
# the world's population. Your task is to download a data set of your choice
# and create 2-5 plots that make use of the techniques from Lesson 3.

# You might use a simple histogram, a boxplot split over a categorical variable,
# or a frequency polygon. The choice is yours!

#import library
library(readxl)
suicide_0_14 <- read_excel("Suicide 0-14 all indicator file WHO version 2 june 2010.xlsx", sheet = "Data")
names(suicide_0_14)[-1] <- as.character(as.numeric(names(suicide_0_14)[-1]))

suicide_15_29 <- read_excel("Suicide 15-29 all indicator file WHO version 2 june 2010.xlsx", sheet = "Data")
names(suicide_15_29)[-1] <- as.character(as.numeric(names(suicide_15_29)[-1]))

suicide_30_44 <- read_excel("Suicide 30-44 all indicator file WHO version 2 june 2010.xlsx", sheet = "Data")
names(suicide_30_44)[-1] <- as.character(as.numeric(names(suicide_30_44)[-1]))

suicide_45_59 <- read_excel("Suicide 45-59 all indicator file WHO version 2 june 2010.xlsx", sheet = "Data")
names(suicide_45_59)[-1] <- as.character(as.numeric(names(suicide_45_59)[-1]))

suicide_60_plus <- read_excel("Suicide 60+ all indicator file WHO version 2 june 2010.xlsx", sheet = "Data")
names(suicide_60_plus)[-1] <- as.character(as.numeric(names(suicide_60_plus)[-1]))
```

```{r}

p1 = qplot(x = subset(suicide_45_59, suicide_45_59[1] == 'United States'),
           data = suicide_0_14)

#p1 = qplot(x = carat, data = suicide_0_14,
#      binwidth = 0.1, geom = 'freqpoly') + 
#      scale_x_continuous(breaks = seq(0,5,by=0.1), limits = c(0,3)) +
#      scale_y_continuous(breaks = seq(0,15000,by=1000))
```

```{r}
US_45_59 <- subset(suicide_45_59, suicide_45_59[1] == 'United States')[-1]
US_45_59
```

```{r Exploring Your Friends' Birthdays}
birthdays <- read.csv(file="birthdaysExample.csv")
# library(lubridate)
birthdays$dates <- mdy(birthdays$dates)
p1 = qplot(x = carat, data = suicide_0_14,
      binwidth = 0.1, geom = 'freqpoly') + 
      scale_x_continuous(breaks = seq(0,5,by=0.1), limits = c(0,3)) +
      scale_y_continuous(breaks = seq(0,15000,by=1000))

# How many people share your birthday? Do you know them?
# (Reserve time with them or save money to buy them a gift!)

# Which month contains the most number of birthdays?

# How many birthdays are in each month?

# Which day of the year has the most number of birthdays?

# Do you have at least 365 friends that have birthdays on everyday
# of the year?
```

Full Details:

# Your task is to investigate the distribution of your friends'
# birth months and days.

# Here some questions you could answer, and we hope you think of others.

# **********************************************************************

# How many people share your birthday? Do you know them?
# (Reserve time with them or save money to buy them a gift!)

# Which month contains the most number of birthdays?

# How many birthdays are in each month?

# Which day of the year has the most number of birthdays?

# Do you have at least 365 friends that have birthdays on everyday
# of the year?

# **********************************************************************

# You will need to do some data munging and additional research to
# complete this task. This task won't be easy, and you may encounter some
# unexpected challenges along the way. We hope you learn a lot from it though.

# You can expect to spend 30 min or more on this task depending if you
# use the provided data or obtain your personal data. We also encourage you
# to use the lubridate package for working with dates. Read over the documentation
# in RStudio and search for examples online if you need help.

# You'll need to export your Facebooks friends' birthdays to a csv file.
# You may need to create a calendar of your Facebook friends’ birthdays
# in a program like Outlook or Gmail and then export the calendar as a
# csv file.

# Once you load the data into R Studio, you can use the strptime() function
# to extract the birth months and birth days. We recommend looking up the
# documentation for the function and finding examples online.

# We've included some links in the Instructor Notes to help get you started.

# Once you've completed your investigation, create a post in the discussions
# that includes:
#   1. any questions you answered, your observations, and summary statistics
#   2. snippets of code that created the plots
#   3. links to the images of your plots

# You can save images by using the ggsave() command.
# ggsave() will save the last plot created.
# For example...
#                  qplot(x = price, data = diamonds)
#                  ggsave('priceHistogram.png')

# ggsave currently recognises the extensions eps/ps, tex (pictex),
# pdf, jpeg, tiff, png, bmp, svg and wmf (windows only).

# Copy and paste all of the code that you used for
# your investigation below the line. Submit it when you are ready.
# ===============================================================================