Lesson 4
========================================================

***

### Scatterplots and Perceived Audience Size
Notes:
there are distinct lines at common guess breakpoints like 50 or 100. Also, you can see that a lot of the data is resting at the bottom, which means they are underestimating the people that see.
***

### Scatterplots
Notes:

```{r Scatterplots}
# set working directory
setwd('/Users/charlesdellinger/Documents/GitHub/Udacity/Exploratory\ Data\ Analysis/eda-course-materials/lesson3')
library(ggplot2)
# pf <- read.csv('pseudo_facebook.tsv', sep='\t')
# could also do:
pf <- read.delim("pseudo_facebook.tsv")

```
```{r Scatterplots - cont}
qplot(x = age, y = friend_count, data = pf)
# can also be written as:
# qplot(age, friend_count, data = pf)
# qplot automatically assumes qpolt(x, y, data=)
```

***

#### What are some things that you notice right away?
Response:
There are a lot of advanced ages where I wouldn't expect people to be alive or use computers. There are a lot of points in the younger user area and they have a lot of friends.
***

### ggplot Syntax
Notes:

```{r ggplot Syntax}
ggplot(aes(x = age, y = friend_count), data = pf) +
  geom_point() + xlim(13,90)

summary(pf$age)
```

***

### Overplotting
Notes:
can set transparency using alpha. alpha = 1/20 means it takes 20 points to equal one regular point.
age is set to years, even though its really continuous. using jitter spaces it out by adding noise and makes a more dispersed distribution.
```{r Overplotting}
ggplot(aes(x = age, y = friend_count), data = pf) +
  geom_point(alpha = 1/20) + xlim(13,90)
ggplot(aes(x = age, y = friend_count), data = pf) +
  geom_jitter(alpha = 1/20) + xlim(13,90)
```

#### What do you notice in the plot?
Response:
the alpha makes the friend counts have a low baseline over all ages, which makes the difference seem less extreme.
***

### Coord_trans()
Notes:
quiz is to add sqrt scale to y (friend_count)
```{r Coord_trans()}
# jitter both adds and subtracts, so using jitter gives negative which with
# sqrt gives imaginary numbers
ggplot(aes(x = age, y = friend_count), data = pf) +
  geom_point(alpha = 1/20) + xlim(13,90) +
  coord_trans(y = "sqrt")
# to add jitter, need to use more advanced syntax
ggplot(aes(x = age, y = friend_count), data = pf) +
  geom_point(alpha = 1/20, position = position_jitter(h = 0)) +
  xlim(13,90) +
  coord_trans(y = "sqrt")
```

#### Look up the documentation for coord_trans() and add a layer to the plot that transforms friend_count using the square root function. Create your plot!

```{r}
# not sure why this is separate from above since the directions weren't provided

# jitter both adds and subtracts, so using jitter gives negative which with
# sqrt gives imaginary numbers
ggplot(aes(x = age, y = friend_count), data = pf) +
  geom_point(alpha = 1/20) + xlim(13,90) +
  coord_trans(y = "sqrt")
# to add jitter, need to use more advanced syntax
ggplot(aes(x = age, y = friend_count), data = pf) +
  geom_point(alpha = 1/20, position = position_jitter(h = 0)) +
  xlim(13,90) +
  coord_trans(y = "sqrt")
```

#### What do you notice?

***

### Alpha and Jitter
Notes:

```{r Alpha and Jitter}
# Examine the relationship between
# friendships_initiated (y) and age (x)
# using the ggplot syntax.

# We recommend creating a basic scatter
# plot first to see what the distribution looks like.
# and then adjusting it by adding one layer at a time.

# What are your observations about your final plot?

# Remember to make adjustments to the breaks
# of the x-axis and to use apply alpha and jitter.
qplot(x = age, y = friendships_initiated, data = pf)
ggplot(aes(x = age, y = friendships_initiated), data = pf) +
  geom_point(alpha = 1/20)
#names(pf)
```

***

### Overplotting and Domain Knowledge
Notes:

***

### Conditional Means
Notes:

```{r Conditional Means}
library(dplyr)
# commonly used functions:
#    filter()
#    group_by()
#    mutate()
#    arrange()

age_groups <- group_by(pf, age)
pf.fc_by_age <- summarise(age_groups,
          friend_count_mean = mean(friend_count),
          friend_count_median = median(friend_count),
          n = n())

pf.fc_by_age <- arrange(pf.fc_by_age, age)
  
head(pf.fc_by_age)

# another way to do the same thing
pf.fc_by_age <- pf %>%
  group_by(age) %>%
  summarise(friend_count_mean = mean(friend_count),
            friend_count_median = median(friend_count),
            n = n()) %>%
  arrange(age)

head(pf.fc_by_age)
```


Create your plot!

```{r Conditional Means Plot}
ggplot(aes(x = age, y = friend_count_mean), data = pf.fc_by_age) +
  geom_line() + xlim(13,80) + coord_trans(y = "log10")
```

***

### Overlaying Summaries with Raw Data
Notes:

```{r Overlaying Summaries with Raw Data}
ggplot(aes(x = age, y = friend_count), data = pf) +
  xlim(13,80) + 
  geom_point(alpha = 0.05,
             position = position_jitter(h = 0),
             color = 'orange') +
  coord_trans(y = "sqrt") + 
  geom_line(stat = 'summary', fun.y = mean) +
  geom_line(stat = 'summary', fun.y = quantile,
            fun.args=list(probs=0.1),
            linetype = 2, color = 'blue') +
    geom_line(stat = 'summary', fun.y = quantile,
            fun.args=list(probs=0.5),
            linetype = 1, color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile,
            fun.args=list(probs=0.9),
            linetype = 2, color = 'blue')

ggplot(aes(x = age, y = friend_count), data = pf) +
  coord_cartesian(xlim = c(13, 70), ylim = c(0, 1000)) +
  geom_point(alpha = 0.05,
             position = position_jitter(h = 0),
             color = 'orange') +
  coord_trans(y = "sqrt") + 
  geom_line(stat = 'summary', fun.y = mean) +
  geom_line(stat = 'summary', fun.y = quantile,
            fun.args=list(probs=0.1),
            linetype = 2, color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile,
            fun.args=list(probs=0.5),
            linetype = 1, color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile,
            fun.args=list(probs=0.9),
            linetype = 2, color = 'blue')
```

#### What are some of your observations of the plot?
Response:

***

### Moira: Histogram Summary and Scatterplot
See the Instructor Notes of this video to download Moira's paper on perceived audience size and to see the final plot.

Notes:

***

### Correlation
Notes:

```{r Correlation}
cor.test(pf$age, pf$friend_count)
# alternative way
with(pf, cor.test(age, friend_count))
```

Look up the documentation for the cor.test function.

What's the correlation between age and friend count? Round to three decimal places.
Response:

***

### Correlation on Subsets
Notes:

```{r Correlation on Subsets}
with(subset(pf, age <= 70), cor.test(age, friend_count))
```

***

### Correlation Methods
Notes:
```{r}
with(subset(pf, age <= 70), cor.test(age, friend_count,
                                    method = 'spearman'))
```
***

## Create Scatterplots
Notes:

```{r}
# Create a scatterplot of likes_received (y)
# vs. www_likes_received (x). Use any of the
# techniques that you've learned so far to
# modify the plot.
ggplot(aes(x = www_likes_received, y = likes_received), data = pf) +
  coord_cartesian(xlim = c(0, 10000), ylim = c(0, 10000)) +
  geom_point(alpha = 1/20) + geom_abline()

# video solution
ggplot(aes(x = www_likes_received, y = likes_received), data = pf) +
  geom_point()
```

***

### Strong Correlations
Notes:

```{r Strong Correlations}
ggplot(aes(x = www_likes_received, y = likes_received), data = pf) +
  geom_point() +
  xlim(0, quantile(pf$www_likes_received, 0.95)) +
  ylim(0, quantile(pf$likes_received, 0.95)) +
  geom_smooth(method = 'lm', color = 'red')
```

What's the correlation betwen the two variables? Include the top 5% of values for the variable in the calculation and round to 3 decimal places.

```{r Correlation Calcuation}
with(pf, cor.test(www_likes_received, likes_received))
```

Response:
the correlation is really high, which is really because one is a superset of another
***

### Moira on Correlation
Notes:
if you measure correlation between variables first, then can determine what you want to keep or not. if you use multiple variables that are highly correlated, its hard to see what's actually driving the result.
***

### More Caution with Correlation
Notes:

```{r More Caution With Correlation}
install.packages('alr3', dependencies = TRUE)
library(alr3)
```

Create your plot!
```{r Temp vs Month}
data(Mitchell)
# basic plot
ggplot(aes(x = Month, y = Temp), data = Mitchell) + geom_point()
# graph month of year data together
ggplot(aes(x = Month %% 12, y = Temp), data = Mitchell) + geom_point()
```

***

### Noisy Scatterplots
a. Take a guess for the correlation coefficient for the scatterplot.
Just the regular month and temp will be terrible, like 0.1
b. What is the actual correlation of the two variables?
(Round to the thousandths place)

```{r Noisy Scatterplots}
with(Mitchell, cor.test(Month, Temp))
with(Mitchell, cor.test(Month %% 12, Temp))
```

***

### Making Sense of Data
Notes:

month <- function(x) {
  x %% 12
}

```{r Making Sense of Data}
#trans_new('month_number', function(x) x %% 12)
ggplot(aes(x = Month, y = Temp), data = Mitchell) + 
  geom_point() + scale_x_continuous(breaks=seq(0,203,12))
```

***

### A New Perspective

What do you notice?
Response:
the temperature values seem to be taking a repeating fluctuation that repeats every year.

Watch the solution video and check out the Instructor Notes!
Notes:
```{r}
ggplot(aes(x=(Month%%12),y=Temp), data=Mitchell)+
  geom_point()
```
***

### Understanding Noise: Age to Age Months
Notes:

```{r Understanding Noise: Age to Age Months}
ggplot(aes(x = age, y = friend_count_mean), data = pf.fc_by_age) +
  geom_line()

head(pf.fc_by_age,10)
pf.fc_by_age[17:19, ]
```

### Create an age_with_months variable

```{r Age with Months}
# Assume the reference date for calculating age is December 31, 2013 and that the age variable gives age in years at the end of 2013.
pf$age_with_months <- pf$age + (1-pf$dob_month/12)
# alternative solution
# pf$age_with_months <- with(pf, age + (1 - dob_month / 12))
```
***

### Age with Months Means

```{r Age with Months Means}
# mean & median friend count and users in each group
age_w_month_groups <- group_by(pf, age_with_months)
pf.fc_by_age_months <- summarise(age_w_month_groups,
          friend_count_mean = mean(friend_count),
          friend_count_median = median(friend_count),
          n = n())

pf.fc_by_age_months <- arrange(pf.fc_by_age_months, age_with_months)
  
head(pf.fc_by_age_months)
# alternative:
age_w_month_groups <- group_by(pf, age_with_months)
pf.fc_by_age_months <- pf %>%
  group_by(age_with_months) %>%
  summarise(friend_count_mean = mean(friend_count),
            friend_count_median = median(friend_count),
            n = n()) %>%
  arrange(age_with_months)
head(pf.fc_by_age_months)
```

Programming Assignment
```{r Programming Assignment}
# Create a new line plot showing friend_count_mean versus the new variable,
# age_with_months. Be sure to use the correct data frame (the one you created
# in the last exercise) AND subset the data to investigate users with ages less
# than 71.
ggplot(aes(x = age_with_months, y = friend_count_mean),
       data = subset(pf.fc_by_age_months, age_with_months < 71)) +
  geom_line()
```

***

### Noise in Conditional Means

```{r Noise in Conditional Means}

```

***

### Smoothing Conditional Means
Notes:

```{r Smoothing Conditional Means}
p1 <- ggplot(aes(x = age, y = friend_count_mean),
             data = subset(pf.fc_by_age, age < 71)) +
  geom_line() +
  geom_smooth()
p2 <- ggplot(aes(x = age_with_months, y = friend_count_mean),
       data = subset(pf.fc_by_age_months, age_with_months < 71)) +
  geom_line() +
  geom_smooth()
p3 <- ggplot(aes(x = round(age/5)*5, y = friend_count),
             data = subset(pf, age < 71)) +
  geom_line(stat = 'summary', fun.y = mean)
library(gridExtra)
grid.arrange(p2, p1, p3, ncol = 1)
```

***

### Which Plot to Choose?
Notes:

***

### Analyzing Two Variables
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

