---
title: "Two Variable Exercises"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# In this problem set, you'll continue
# to explore the diamonds data set.

```{r load data and ggplot library}
data(diamonds)
library(ggplot2)
```
# Your first task is to create a
# scatterplot of price vs x.
# using the ggplot syntax.
```{r}
with(diamonds, summary(price))
with(diamonds, summary(x))
```
```{r}
ggplot(aes(x = x, y = price),
       data = diamonds) +
  geom_point()
```

```{r Correlations}
# What is the correlation between price and x?
with(diamonds, cor.test(price, x))
# What is the correlation between price and y?
with(diamonds, cor.test(price, y))
# What is the correlation between price and z?
with(diamonds, cor.test(price, z))
```

```{r price vs. depth}
# Create a simple scatter plot of price vs depth.
ggplot(aes(x = depth, y = price), data = diamonds) +
  geom_point()
```

```{r Adjustments - price vs. depth}
# Change the code to make the transparency of the
# points to be 1/100 of what they are now and mark
# the x-axis every 2 units. See the instructor notes
# for two hints.
ggplot(aes(x = depth, y = price), data = diamonds) +
  geom_point(alpha = 1/100) +
  scale_x_continuous(breaks=seq(0,80,2))
```

```{r Typical Depth Range}
# Based on the scatterplot of depth vs. price, most of the diamonds are between what values of depth?
with(diamonds, quantile(depth, 0.1))
with(diamonds, quantile(depth, 0.9))
# 60 to 64
```

```{r Correlation - price and depth}
# What's the correlation of depth vs. price?
with(diamonds, cor.test(price, depth))
# Based on the correlation coefficient would you use depth to predict the price of a diamond?
No, the correlation coefficient is too small to show that they trend together.
```

```{r price vs. carat}
# Create a scatterplot of price vs carat
# and omit the top 1% of price and carat
# values.
ggplot(aes(x = carat, y = price),
       data = subset(diamonds, carat < quantile(carat, .99) &
                       price < quantile(price, 0.99))) +
  geom_point(alpha = 1/100)
```

```{r price vs. volume}
# Create a scatterplot of price vs. volume (x * y * z).
# This is a very rough approximation for a diamond's volume.
ggplot(aes(x = x*y*z, y = price),
       data = diamonds) +
  geom_point() +
  xlim(c(0,1000))
# Create a new variable for volume in the diamonds data frame.
# This will be useful in a later exercise.
diamonds$volume <- with(diamonds, x*y*z)
# Don't make any adjustments to the plot just yet.
```
Did you notice some outliers? Some volumes are 0! There's an expensive diamond with a volume near 4000, and a cheaper diamond with a volume near 900.

You can find out how many diamonds have 0 volume by using `count(diamonds$volume == 0)`. The `count()` function comes with the plyr package.

Note: If you ran the `count` function from plyr, you need to run this command in R to unload the plyr package.
```detach("package:plyr", unload=TRUE)```
The plyr package will conflict with the dplyr package in later exercises.

Depending on your investigation, it may or may not be important for you to understand how outliers, like these, came to be in your data.

```{r Correlations on Subsets}
# What's the correlation of price and volume?
# Exclude diamonds that have a volume of 0 or that are greater than or equal to 800.
with(subset(diamonds, (volume > 0 & volume <= 800)),
     cor.test(price, volume))
```
```{r Adjustments - price vs. volume}
# Subset the data to exclude diamonds with a volume
# greater than or equal to 800. Also, exclude diamonds
# with a volume of 0. Adjust the transparency of the
# points and add a linear model to the plot. (See the
# Instructor Notes or look up the documentation of
# geom_smooth() for more details about smoothers.)
p1 <- ggplot(aes(x = volume, y = price),
       data = subset(diamonds, (volume > 0 & volume <= 800))) +
  geom_point(alpha = 1/20)
p1 + geom_smooth(method = "glm", formula = y ~ poly(x,2))
# Do you think this would be a useful model to estimate
# the price of diamonds? Why or why not?

# this line isn't particularly useful. the outliers pull
# the regression line out, so it doesn't arc up as fast 
# as the bulk of the data. There are also clusters at
# certain breakpoints that a simple curve doesn't capture.
```

```{r Mean Price by Clarity}
# Use the function dplyr package
# to create a new data frame containing
# info on diamonds by clarity.

# Name the data frame diamondsByClarity

# The data frame should contain the following
# variables in this order.

#       (1) mean_price
#       (2) median_price
#       (3) min_price
#       (4) max_price
#       (5) n

# where n is the number of diamonds in each
# level of clarity.

data(diamonds)
library(dplyr)

diamondsByClarity <- diamonds %>%
  group_by(clarity) %>%
  summarize(mean_price = mean(price),
            median_price = median(price),
            min_price = min(price),
            max_price = max(price),
            n = n()) %>%
    arrange(clarity)
head(diamondsByClarity)
```

```{r Bar Charts of Mean Price}
# We’ve created summary data frames with the mean price
# by clarity and color. You can run the code in R to
# verify what data is in the variables diamonds_mp_by_clarity
# and diamonds_mp_by_color.

# Your task is to write additional code to create two bar plots
# on one output image using the grid.arrange() function from the package
# gridExtra.

# DO NOT DELETE THE LINES OF CODE BELOW
# ===================================================================
data(diamonds)
library(dplyr)

diamonds_by_clarity <- group_by(diamonds, clarity)
diamonds_mp_by_clarity <- summarise(diamonds_by_clarity, mean_price = mean(price))

diamonds_by_color <- group_by(diamonds, color)
diamonds_mp_by_color <- summarise(diamonds_by_color, mean_price = mean(price))
```
```{r}
# ENTER YOUR CODE BELOW THIS LINE
# ===================================================================
library(gridExtra)

b1 <- ggplot(aes(y = mean_price, x = clarity), data = diamonds_mp_by_clarity) +
  geom_col()
b2 <- ggplot(aes(y = mean_price, x = color), data = diamonds_mp_by_color) +
  geom_col()
grid.arrange(b1, b2, ncol = 1)
```

What do you notice in each of the bar charts for mean price by clarity and mean price by color?

There's an inverse relationship between grades of clarity/color and mean price, which seems backwards from what is expected.

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```