Lesson 3
========================================================

***

### What to Do First?
Notes:
Get the big picture of the data.
Start by loading data set.
***

### Pseudo-Facebook User Data
Notes:

```{r Pseudo-Facebook User Data}
pf <- read.csv("pseudo_facebook.tsv", sep = '\t')
```

***

### Histogram of Users' Birthdays
Notes:

```{r Histogram of Users\' Birthdays}
# install.packages('ggplot2')
library(ggplot2)

names(pf)
qplot(x = dob_day, data = pf) +
  scale_x_continuous(breaks=1:31)
```

***

#### What are some things that you notice about this histogram?
Response:
The first day of the month is much higher. Not all months have 31 days, so that bin should be lower.
***

### Moira's Investigation
Notes:
Perception of audience on social media matches up with who is actually reading.
***

### Estimating Your Audience Size
Notes:

***

#### Think about a time when you posted a specific message or shared a photo on Facebook. What was it?
Response:

#### How many of your friends do you think saw that post?
Response:

#### Think about what percent of your friends on Facebook see any posts or comments that you make in a month. What percent do you think that is?
Response:

***

### Perceived Audience Size
Notes:
people underestimated their audience by about 1/4th.
***
### Faceting
Notes:

```{r Faceting}
ggplot(data = pf, aes(x = dob_day)) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(breaks = 1:31) +
  facet_wrap(~dob_month)
```

#### Let’s take another look at our plot. What stands out to you here?
Response:

***

### Be Skeptical - Outliers and Anomalies
Notes:

***

### Moira's Outlier
Notes:
#### Which case do you think applies to Moira’s outlier?
Response:

***

### Friend Count
Notes:

#### What code would you enter to create a histogram of friend counts?

```{r Friend Count}
#str(pf)
qplot(x = friend_count, data = pf)
```

#### How is this plot similar to Moira's first plot?
Response:
Most of the conuts are squished on the left side, so most of the analysis is conserned with data that is not easily visible sue to scaling for the tail.
***

### Limiting the Axes
Notes:

```{r Limiting the Axes}
qplot(x = friend_count, data = pf, xlim = c(0, 1000))

qplot(x = friend_count, data = pf) +
  scale_x_continuous(limits = c(0, 1000))
```

### Exploring with Bin Width
Notes:

***

### Adjusting the Bin Width
Notes:
```{r}
qplot(x = friend_count, data = pf, binwidth = 25) +
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50))
```

### Faceting Friend Count
```{r Faceting Friend Count}
# What code would you add to create a facet the histogram by gender?
# Add it to the code below.
qplot(x = friend_count, data = pf, binwidth = 10) +
  scale_x_continuous(limits = c(0, 1000),
                     breaks = seq(0, 1000, 50)) +
  facet_grid(gender ~.)
```

***

### Omitting NA Values
Notes:

```{r Omitting NA Values}
qplot(x = friend_count, data = subset(pf, !is.na(gender)), binwidth = 10) +
  scale_x_continuous(limits = c(0, 1000),
                     breaks = seq(0, 1000, 50)) +
  facet_grid(gender ~.)
```

***

### Statistics 'by' Gender
Notes:

```{r Statistics \'by\' Gender}
table(pf$gender)
by(pf$friend_count, pf$gender, summary)
```

#### Who on average has more friends: men or women?
Response: female

#### What's the difference between the median friend count for women and men?
Response: 22

#### Why would the median be a better measure than the mean?
Response: there is are huge maximums that might be skewing the mean

***

### Tenure
Notes:

```{r Tenure}
qplot(x = tenure, data = pf, binwidth = 30,
      color = I('black'), fill = I('#099DD9'))
```

***

#### How would you create a histogram of tenure by year?

```{r Tenure Histogram by Year}
qplot(x = tenure/365, data = pf, binwidth = 0.25,
      color = I('black'), fill = I('#F79420')) +
  scale_x_continuous(breaks = seq(1,7,1), limits=c(1,7))
```

***

### Labeling Plots
Notes:

```{r Labeling Plots}
qplot(x = tenure/365, data = pf, binwidth = 0.25,
      xlab = 'Number of years using Facebook',
      ylab = 'Number of users in sample',
      color = I('black'), fill = I('#F79420')) +
  scale_x_continuous(breaks = seq(1,7,1), limits=c(1,7))
```

***

### User Ages
Notes:

```{r User Ages}
qplot(x = age, data = pf, binwidth = 5,
      xlab = 'Age of Facebook user',
      ylab = 'Number of users in sample',
      color = I('black'), fill = I('#F79420')) #+
  #scale_x_continuous(breaks = seq(0,60,5), limits=c(0,65))

qplot(x = age, data = pf, binwidth = 5,
      xlab = 'Age of Facebook user',
      ylab = 'Number of users in sample',
      color = I('black'), fill = I('#F79420')) +
  scale_x_continuous(breaks = seq(0,60,5), limits=c(0,65))

qplot(x = age, data = pf, binwidth = 1,
      xlab = 'Age of Facebook user',
      ylab = 'Number of users in sample',
      color = I('black'), fill = I('#F79420')) +
  scale_x_continuous(breaks = seq(12,30,1), limits=c(12,35))
```
```{r age example from video}
# example from video
qplot(x = age, data = pf, binwidth = 1,
      xlab = 'Age of Facebook user',
      ylab = 'Number of users in sample',
      color = I('black'), fill = I('#5760AB')) +
  scale_x_continuous(breaks = seq(0,113,5))
```
#### What do you notice?
Response:
Largest amount of users are in their early 20s. No one is registered below the minimum age of 13 required to register. There are some extreme ages (>100 years old) that don't seem real.
***

### The Spread of Memes
Notes:
Individual to individual infomraiton transfer. Memes replicate through the social network. They tend to recur over time.
***

### Lada's Money Bag Meme
Notes:
The plot makes it look likes it disappears. When zooming in using a log scale, you can see that the meme persists in low numbers after the huge drop off.
***

### Transforming Data
Notes:
```{r Transforming Data}
#qplot(x = friend_count, data = pf)


summary(pf$friend_count)
summary(log10(pf$friend_count + 1))
summary(sqrt(pf$friend_count))

library(gridExtra)
# define individual plots
p1 <- qplot(x = friend_count, data = pf)
p2 <- qplot(x = log10(friend_count+1), data = pf)
p3 <- qplot(x = sqrt(friend_count), data = pf)
# arrange plots in grid
grid.arrange(p1, p2, p3, ncol=1)
```
```{r Transforming Data - Alternate Answer}
p1 <- ggplot(aes(x = friend_count), data = pf) + geom_histogram()
p2 <- p1 + scale_x_log10()
p3 <- p1 + scale_x_sqrt()
grid.arrange(p1, p2, p3, ncol=1)
```
***

### Add a Scaling Layer
Notes:
adding a layer converts the scale of the axis. For instance, graphing the histogram for log10(x) graphs the log values vs applying a layer which uses units 1...1000, etc.
```{r Add a Scaling Layer}
# see above scaling layers
```

***


### Frequency Polygons

```{r Frequency Polygons}
qplot(x = friend_count, data = subset(pf, !is.na(gender)),
      binwidth = 10) +
  scale_x_continuous(lim = c(0, 1000), breaks = seq(0, 1000, 50)) + 
  facet_wrap(~gender)
# use overlapping frequency polygon
qplot(x = friend_count, data = subset(pf, !is.na(gender)),
      binwidth = 10, geom = 'freqpoly', color = gender) +
  scale_x_continuous(lim = c(0, 1000), breaks = seq(0, 1000, 50))

qplot(x = friend_count, y = ..count../sum(..count..),
      data = subset(pf, !is.na(gender)),
      xlab = 'Friend Count',
      ylab = 'Proportion of Users with that friend count',
      binwidth = 10, geom = 'freqpoly', color = gender) +
  scale_x_continuous(lim = c(0, 1000), breaks = seq(0, 1000, 50))
```
```{r}
qplot(x = friend_count, y = ..count../sum(..count..),
      data = subset(pf, !is.na(gender)),
      xlab = 'Friend Count',
      ylab = 'Proportion of Users with that friend count',
      binwidth = 10, geom = 'freqpoly', color = gender) +
  scale_x_continuous(lim = c(500, 1000), breaks = seq(0, 1000, 50))

qplot(x = friend_count, y = ..count../sum(..count..),
      data = subset(pf, !is.na(gender)),
      xlab = 'Friend Count',
      ylab = 'Proportion of Users with that friend count',
      binwidth = 10, geom = 'freqpoly', color = gender) +
      scale_x_log10(limits = c(500, 1000))
```
***

### Likes on the Web
Notes:

```{r Likes on the Web}
# names(pf)
qplot(x = www_likes, data = subset(pf, !is.na(gender)),
      geom = 'freqpoly', color = gender) +
  scale_x_continuous() +
  scale_x_log10()
```
```{r}
by(pf$www_likes, pf$gender, sum)
```

***

### Box Plots
Notes:
x <- categorical data
y <- continuous data

```{r Box Plots}
qplot(x = gender, y = friend_count,
      data = subset(pf, !is.na(gender)),
      geom = 'boxplot') +
  scale_y_log10()
```

#### Adjust the code to focus on users who have friend counts between 0 and 1000.

```{r}
qplot(x = gender, y = friend_count,
      data = subset(pf, !is.na(gender)),
      geom = 'boxplot') +
  scale_y_continuous(limits = c(0,1000)) +
  scale_y_log10()

qplot(x = gender, y = friend_count,
      data = subset(pf, !is.na(gender)),
      geom = 'boxplot', ylim = c(0,1000))

qplot(x = gender, y = friend_count,
      data = subset(pf, !is.na(gender)),
      geom = 'boxplot') + 
  coord_cartesian(ylim = c(0,1000))
```

***

### Box Plots, Quartiles, and Friendships
Notes:

```{r Box Plots, Quartiles, and Friendships}
qplot(x = gender, y = friend_count,
      data = subset(pf, !is.na(gender)),
      geom = 'boxplot') + 
  coord_cartesian(ylim = c(0,250))

by(pf$friend_count, pf$gender, summary)
```

#### On average, who initiated more friendships in our sample: men or women?
Response:
```{r}
qplot(x = gender, y = friendships_initiated,
      data = subset(pf, !is.na(gender)),
      geom = 'boxplot') + 
  coord_cartesian(ylim = c(0,150))

by(pf$friendships_initiated, pf$gender, summary)
```
#### Write about some ways that you can verify your answer.
Response:
```{r Friend Requests by Gender}
by(pf$friendships_initiated, pf$gender, summary)
```

Response:
You can look at the numeric summary to double check the boxplot.
***

### Getting Logical
Notes:

```{r Getting Logical}
summary(pf$mobile_likes)

summary(pf$mobile_likes > 0)

pf$mobile_check_in <- NA
pf$mobile_check_in <- ifelse(pf$mobile_likes > 0,1,0)
pf$mobile_check_in <- factor(pf$mobile_check_in)
summary(pf$mobile_check_in)
```

Response:
```{r}
sum(pf$mobile_check_in == 1)/length(pf$mobile_check_in == 1)
```
***

### Analyzing One Variable
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!